Grammar-Based Concept Alignment for Domain-Specific Machine Translation

### 2021-05-16 paper/paper.tex
squeezed in sentence about commercial projects and other las minute changes

% Author information can be set in various styles:
% For several authors from the same institution:
% \author{Author 1 \and ... \and Author n \\
%         Address line \\ ... \\ Address line}
% if the names do not fit well on one line use
%         Author 1 \\ {\bf Author 2} \\ ... \\ {\bf Author n} \\
% For authors from different institutions:
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \And  ... \And
%         Author n \\ Address line \\ ... \\ Address line}
% To start a seperate ``row'' of authors use \AND, as in
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \AND
%         Author 2 \\ Address line \\ ... \\ Address line \And
%         Author 3 \\ Address line \\ ... \\ Address line}
\author{First Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  \texttt{email@domain} \\\And
  Second Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  \texttt{email@domain} \\}

@@

\begin{figure*}[h]
  \centering
  \scriptsize
  \begin{subfigure}{.30  \textwidth}
    \setlength{\unitlength}{0.14mm}
\begin{picture}(257.0,70.0)
  \put(0.0,0.0){She}
  \put(46.0,0.0){studies}
  \put(119.0,0.0){consistently}
  \put(0.0,15.0){{\tiny PRON}}
  \put(46.0,15.0){{\tiny VERB}}
  \put(119.0,15.0){{\tiny ADV}}
  \put(33.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
  \put(13.26086956521739,35.0){\vector(0,-1){5.0}}
  \put(18.0,49.66666666666667){{\tiny nsubj}}
  \put(102.5,30.0){\oval(68.89041095890411,33.333333333333336)[t]}
  \put(136.94520547945206,35.0){\vector(0,-1){5.0}}
  \put(87.5,49.66666666666667){{\tiny advmod}}
  \put(61.0,70.0){\vector(0,-1){40.0}}
  \put(66.0,60.0){{\tiny root}}
\end{picture}
\\ \\
\begin{picture}(259.0,90.0)
  \put(0.0,0.0){Lei}
  \put(46.0,0.0){studia}
  \put(110.0,0.0){con}
  \put(147.0,0.0){costanza}
  \put(0.0,15.0){{\tiny PRON}}
  \put(46.0,15.0){{\tiny NOUN}}
  \put(110.0,15.0){{\tiny ADP}}
  \put(147.0,15.0){{\tiny NOUN}}
  \put(33.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
  \put(13.26086956521739,35.0){\vector(0,-1){5.0}}
  \put(18.0,49.66666666666667){{\tiny nsubj}}
  \put(138.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
  \put(124.05405405405405,35.0){\vector(0,-1){5.0}}
  \put(123.5,49.66666666666667){{\tiny case}}
  \put(116.5,30.0){\oval(98.02970297029702,66.66666666666667)[t]}
  \put(165.5148514851485,35.0){\vector(0,-1){5.0}}
  \put(101.5,66.33333333333334){{\tiny nmod}}
  \put(61.0,90.0){\vector(0,-1){60.0}}
  \put(66.0,80.0){{\tiny root}}
\end{picture}
  \end{subfigure}
  \begin{subfigure}{.30  \textwidth}
    \begin{verbatim}

@@

1 Lei PRON 2 nsubj
2 studia VERB 0 root
3 con ADP 4 case
4 costanza NOUN 2 obl
    \end{verbatim}
  \end{subfigure}
  \begin{subfigure}{.30 \textwidth}
    \begin{verbatim}

@@

2 studia VERB 0 root
  1 Lei PRON 2 nsubj
  4 costanza NOUN 2 obl
    3 con ADP 4 case
    \end{verbatim}
  \end{subfigure}
  \caption{The graphical, simplified CoNNL-U and sorted rose tree representation of a pair of UD sentences. With the default criteria, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}$\rangle$}. 
\end{figure*}

@@

\subsubsection{Results on manually annotated treebanks} 
% table 1
In Table \ref{pud_fast}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks.

% details
Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files except with regards to tokenization. 
On the other hand, the relatively large size of the PUD treebanks makes it possible also to train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison. 
In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured only to extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases. 
This explains CE's seemingly low recall.
To get a better idea, Table \ref{pud_fast} can be compared with Table \ref{raw_fast}, which summarizes the results of an experiment where no constraints are placed on the size of the extracted alignments.

@@

% test corpus construction
The idea is to automatically translate a set of English sentences to Italian and Swedish, ask native speakers of the target languages to produce a set of reference translations, and compare them to the original machine-generated ones by computing BLEU scores.
Due to the small size of the datasets and the consequently low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell rather than trying to parse arbitrary sentences from other course plans. 
In order to do that, we make use of GF's random AST generation functionality but at the same time manually select semantically plausible sentences to facilitate the task of the human translators.
The results of this process are two small testing corpora, one for the DMI and one for the CSE corpus, each consisting of 50 English sentences.

% reference translations
Reference translations are obtained by asking participants to compare the original English sentences to their automatically translated counterparts and correct the latter by making the minimal changes necessary to obtain a set of grammatically and semantically correct translations. 
This is important as, if the reference translations are obtained independently from the automatic ones, BLEU scores can easily become misleading.

@@

% cool alignment errors
Some errors regarding the extracted concepts are also interesting to analyze: the alignment $\langle$\textit{``class'', ``classe''}$\rangle$, for instance, causes the sentence \textit{``I will attend the class''} to be (incorrectly) translated as \textit{``io seguirò la classe''} instead of \textit{``io seguirò la lezione''} even though the correspondence is in fact valid in most of the numerous contexts in which (within the same domain!) \textit{``class''} is not to be intended as a synonym of \textit{``lesson''}.

@@

% sw
The tangible fruits of this work are a Haskell library and a number of executables offering a user friendly interface to perform CE, lexicon generation and various kinds of evaluations. 
The source code, including a preliminary implementation of CP, is available on GitHub\footnote{\url{github.com/harisont/concept-alignment}}.

>>>>

% Author information can be set in various styles:
% For several authors from the same institution:
% \author{Author 1 \and ... \and Author n \\
%         Address line \\ ... \\ Address line}
% if the names do not fit well on one line use
%         Author 1 \\ {\bf Author 2} \\ ... \\ {\bf Author n} \\
% For authors from different institutions:
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \And  ... \And
%         Author n \\ Address line \\ ... \\ Address line}
% To start a seperate ``row'' of authors use \AND, as in
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \AND
%         Author 2 \\ Address line \\ ... \\ Address line \And
%         Author 3 \\ Address line \\ ... \\ Address line}
\author{Arianna Masciolini \\
  Digital Grammars \\
  \texttt{arianna@digitalgrammars.com} \\\And
  Aarne Ranta \\
  Department of Computer Science and Engineering, \\
  University of Gothenburg \\
  Digital Grammars \\
  \texttt{aarne.ranta@cse.gu.se} \\}

@@

\begin{figure*}[h]
  \centering
  \scriptsize
  \begin{subfigure}{.32  \textwidth}
    \setlength{\unitlength}{0.14mm}
\begin{picture}(257.0,70.0)
  \put(0.0,0.0){She}
  \put(46.0,0.0){studies}
  \put(119.0,0.0){consistently}
  \put(0.0,15.0){{\tiny PRON}}
  \put(46.0,15.0){{\tiny VERB}}
  \put(119.0,15.0){{\tiny ADV}}
  \put(33.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
  \put(13.26086956521739,35.0){\vector(0,-1){5.0}}
  \put(18.0,49.66666666666667){{\tiny nsubj}}
  \put(102.5,30.0){\oval(68.89041095890411,33.333333333333336)[t]}
  \put(136.94520547945206,35.0){\vector(0,-1){5.0}}
  \put(87.5,49.66666666666667){{\tiny advmod}}
  \put(61.0,70.0){\vector(0,-1){40.0}}
  \put(66.0,60.0){{\tiny root}}
\end{picture}
\\ \\
\begin{picture}(259.0,90.0)
  \put(0.0,0.0){Lei}
  \put(46.0,0.0){studia}
  \put(110.0,0.0){con}
  \put(147.0,0.0){costanza}
  \put(0.0,15.0){{\tiny PRON}}
  \put(46.0,15.0){{\tiny VERB}}
  \put(110.0,15.0){{\tiny ADP}}
  \put(147.0,15.0){{\tiny NOUN}}
  \put(33.0,30.0){\oval(39.47826086956522,33.333333333333336)[t]}
  \put(13.26086956521739,35.0){\vector(0,-1){5.0}}
  \put(18.0,49.66666666666667){{\tiny nsubj}}
  \put(138.5,30.0){\oval(28.89189189189189,33.333333333333336)[t]}
  \put(124.05405405405405,35.0){\vector(0,-1){5.0}}
  \put(123.5,49.66666666666667){{\tiny case}}
  \put(116.5,30.0){\oval(98.02970297029702,66.66666666666667)[t]}
  \put(165.5148514851485,35.0){\vector(0,-1){5.0}}
  \put(101.5,66.33333333333334){{\tiny nmod}}
  \put(61.0,90.0){\vector(0,-1){60.0}}
  \put(66.0,80.0){{\tiny root}}
\end{picture}
  \end{subfigure}
  \begin{subfigure}{.32  \textwidth}
    \begin{verbatim}

@@

1 Lei PRON 2 nsubj
2 studia VERB 0 root
3 con ADP 4 case
4 costanza NOUN 2 obl
    \end{verbatim}
  \end{subfigure}
  \begin{subfigure}{.32 \textwidth}
    \begin{verbatim}

@@

2 studia VERB 0 root
  1 Lei PRON 2 nsubj
  4 costanza NOUN 2 obl
    3 con ADP 4 case
    \end{verbatim}
  \end{subfigure}
  \caption{The graphical, simplified CoNNL-U and sorted rose tree representation of a pair of UD sentences. With the default criteria, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}$\rangle$.}
\end{figure*}

@@

\subsubsection{Results on manually annotated treebanks} 
% table 1
In Table \ref{pud_fast}, we compare the results obtained with our grammar-based module to those obtained statistically on the PUD treebanks. 
Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files except with regards to tokenization. 
On the other hand, the relatively large size of the PUD treebanks makes it possible also to train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison. 
In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured only to extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases. 
This explains CE's seemingly low recall.
To get a better idea, Table \ref{pud_fast} can be compared with Table \ref{raw_fast}, which summarizes the results of an experiment where no constraints are placed on the size of the extracted alignments.

@@

% test corpus construction
The idea is to automatically translate a set of English sentences to Italian and Swedish, ask native speakers of the target languages to produce a set of reference translations, and compare them to the original machine-generated ones by computing BLEU scores.
Due to the small size of the datasets and the consequently low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell rather than trying to parse arbitrary sentences from other course plans. 
In order to do that, we make use of GF's random AST generation functionality but at the same time manually select semantically plausible sentences to facilitate the task of the human translators.
The results of this process are two small testing corpora, one for the DMI and one for the CSE corpus, each consisting of 50 English sentences. 
Reference translations are obtained by asking participants to compare the original English sentences to their automatically translated counterparts and correct the latter with the minimal changes necessary to obtain a set of grammatically and semantically correct translations. 
This is important as, if the reference translations are obtained independently, BLEU scores can easily become misleading.

@@

% cool alignment errors
Some errors regarding the extracted concepts are also interesting to analyze: the alignment $\langle$\textit{class, classe}$\rangle$, for instance, causes the sentence \textit{``I will attend the class''} to be (incorrectly) translated as \textit{``io seguirò la classe''} instead of \textit{``io seguirò la lezione''} even though the correspondence is in fact valid in most contexts in which (within the same domain!) \textit{``class''} is not to be intended as a synonym of \textit{``lesson''} but as teaching group.

@@

% sw
The tangible fruits of this work are a Haskell library and a number of executables offering a user friendly interface to perform CE, lexicon generation and various kinds of evaluations. 
The source code, including a preliminary implementation of CP, is available on GitHub\footnote{link hidden at review time}. % TODO: github.com/harisont/concept-alignment
The software has already been used to analyse customer-provided data in two commercial projects. % TODO: at Digital Grammars.
